---
layout: 12pt,review
linenumbers: true
title: "Supplement for: "Pleistocene-aged stone artefacts from Jerimalai, East Timor: Long term conservatism in early modern human technology in island Southeast Asia""
author: 
  - name: Ben Marwick
    affiliation: uw
    email: bmarwick@uw.edu
    footnote: Corresponding author

address: 
  - code: uw
    address: | 
      Department of Anthropology, 
      Denny Hall M32, Box 353100,  
      University of Washington,
      Seattle, WA 98195-3100 USA 

output: html_document
---

```{r setup, include=FALSE, eval=FALSE}
require(knitr)
require(rmarkdown)
require(ggplot2)
require(reshape2)
require(plyr)
require(dplyr)
require(vcd)
require(rjags)
require(coda)
require(BEST)
require(lattice)
require(xtable)
require(data.table)

opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""))
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

opts_chunk$set(fig.width=4, fig.height=3, echo=FALSE)
theme_set(theme_bw(base_size=12))
toggle = "hide" # results argument

```

# Introduction

This document contains the R code to accompany the analysis of stone artefacts from Jerimalai rockshelter reported in Marwick et al. "Pleistocene-aged stone artefacts from Jerimalai, East Timor: Long term conservatism in early modern human technology in island Southeast Asia". This document is organised with the same structure as the manuscript and will produce all of the quantitative results, data visualisations and tables found in the manuscript. This document is an R markdown document, which means that it is a plain text file that contains marked-up text (such as this paragraph), and chunks of R code in between the text. This file may be viewed in any text editor, and executed to run the code with R v3.2.0. The purpose of providing this supplement is to enable the reader to more thoroughly evaluate the reliability of our work through close inspection the quantitiative methods we used in the paper, and to enable computational reproduciblity of our methods to facailite their application to other research projects. Note that the code presented here has been developed and tested specifically for this analysis and is not yet intended as a general purpose tool. 

## Code

The exact version of the code that produced the data and figures in the published paper is archived at [http://dx.doi.org/10.6084/m9.figshare.985406](http://dx.doi.org/10.6084/m9.figshare.985406). The development version of this code can be found at [https://github.com/benmarwick/Pleistocene-aged-stone-artefacts-from-Jerimalai--East-Timor](https://github.com/benmarwick/Pleistocene-aged-stone-artefacts-from-Jerimalai--East-Timor). The code in the development version may have changed since publication and may not produce exactly the same output as found in the published paper. 
 
The code should run on a typical personal computer (Windows/OSX/Linux) that can run R and can install R packages from the internet. The specific software dependencies for the code included here are listed in the DESCRIPTION file of the R package that this supplement is contained in (JerimalaiStoneArtefacts).  

## Data


```{r load_data}

# load data
 
# dates
dates <- read.csv("data/Jeremalai_dates.csv", as.is = TRUE)

# complete flake data
flakes <- read.csv("data/JB_Chert_Flakes_and_Retouch.csv")

# core type data
core_types <- read.csv("data/Jerimalai_cores_techno_metrics.csv")

# spit depths
depths <- read.csv("data/Jeremalai_spit_depths.csv")

# sediment volumes
vols <- read.csv("data/Artefact densities with soil volumes Sq B.csv", skip = 1)

# all artefacts
all <- read.csv("data/Jerimalai_All_Artefacts.csv")

# techno types
cores <- read.csv("data/Jerimalai_tech_table_cores.csv")
types <-  na.omit(read.csv("data/Jerimalai_tech_table_types.csv"))
retouch <- read.csv("data/Jerimalai_tech_table_retouch.csv")
features <- read.csv("data/Jerimalai_tech_table_features.csv")
ground <- read.csv("data/Jerimalai_tech_table_ground.csv")

# retouch indices
retouch_indices <- read.csv("data/Jerimalai_retouch_indices.csv")
```


# Chronology of the excavated deposit



```{r chronology}

# plot dates and label with phases

dates$cal_age_min <- as.numeric(gsub("\\D", "", dates$cal_age_min))
dates$cal_age_max <- as.numeric(gsub("\\D", "", dates$cal_age_max))
dates$mid <- with(dates, cal_age_min + ((cal_age_max - cal_age_min) /2) )
ggplot(dates, aes(mid, depth_bs)) +
  geom_point() +
  scale_y_reverse() + 
  geom_errorbarh(aes(xmin=cal_age_min,xmax=cal_age_max,  height = 0)) +
  theme_minimal() +
  #geom_smooth(span = 0.19, se = FALSE) +
  ylab("depth below surface (m)") +
  xlab("calibrated age (BP)") +
  geom_vline(xintercept = 42000, colour = 'grey') +
  annotate("text", x = 40000, y = 0.25, label = "Phase I", angle = 90) + 
  geom_vline(xintercept = 38000, colour = 'grey') +
  geom_vline(xintercept = 17000, colour = 'grey') +
  annotate("text", x = 25000, y = 0.25, label = "Phase II", angle = 90) + 
  geom_vline(xintercept = 9000, colour = 'grey') +
  annotate("text", x = 14000, y = 0.25, label = "Phase III", angle = 90) + 
  geom_vline(xintercept = 5500, colour = 'grey') +
  annotate("text", x = 7000, y = 0.25, label = "Phase IV", angle = 90) +
  geom_vline(xintercept = 5500,colour = 'grey') +
  geom_vline(xintercept = 0, colour = 'grey') +
  annotate("text", x = 3500, y = 0.25, label = "Phase V", angle = 90)
  theme( # remove the vertical grid lines
    panel.grid.major.x = element_blank() ,
    # explicitly set the horizontal lines (or they will disappear too)
    panel.grid.major.y = element_blank()  
  )

# From Science paper - four chonological divisions
# The chronological divisions are: 
# Phase I (spits 69-50) dated ~42 to 38 ka; 
# Phase II (spits 49-40) # dated between ~ 17 to 9 ka, 
# Phase III (spits 39-21) dated between 6.5 and 5.5 ka, 
# Phase IV (spits 20-3) dated ~5.5 - modern.

# I have chosen not to omit 17-38 ky BP here 
 
ggsave("figures/Jeremalai-dates.png")
```

# Results: Raw materials

# Results: Discard rates

# Results: Artefact taphonomy 

# Results: Metric and technological characteristics of cores and unretouched flakes 

# Results: Core technology

# Results: Retouched artefacts 

# Results: Technological types

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
